Resource-ratio theory in metaecosystems

### 2013-11-05 ms_resource_ratio.tex
First upload of files

\documentclass[12pt]{paper}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{lineno}
\usepackage[style=numeric-comp,url=false,doi=false,isbn=false,maxnames=2,maxbibnames=2,firstinits,sorting=none]{biblatex}
\bibliography{/Users/DGravel/Documents/library.bib}

@@

%\title{Resource-ratio theory in metaecosystems}

\author{Dominique Gravel}

\maketitle\onehalfspacing\linenumbers

@@

@@

Paragraph 1: contribution of the resource-ratio theory to the understanding of coexistence

Paragraph 2: resource-ratio theory is essentially spatial, but has yet to integrate recent developments of metacommunity theory. Only a few amounts of papers deals with spatial exchanges of nutrients and organisms (Klausmeier; Ryabov and Blasius). These studies concern the vertical distribution of biomass in the water column and do not provide a mechanistic understanding of the impacts of dispersal on coexistence.

Paragraph 3: objectives and approach

@@

%-----------------%
\begin{equation}
	\label{e:bnet}
	\frac{dN_{jx}}{dt}=S_{jx}-eN_{jx}-\sum_{i=1}^{s}q_{ij}G_{ix}N_{jx}+\sum_{i=1}^{s}r(1-\phi)D_{ix}-\Delta_{Nx}
\end{equation}
%-----------------%

@@

The conditions for coexistence of primary producer species competing for two resources is best understood using a graphical representation of the model (Tilman, 1982). We first illustrate and derive tools to interpret the equilibrium state of the system with a single resident primary producer species and detritus mineralization at a single locality. We introduce dispersal for each compartment at the following section.

@@

where $\overrightarrow{u'}_{x}$ is an altered nutrient supply vector. The consumption vector therefore points toward a net nutrient supply point $S'_{x}$, which his new location is given by:

@@

The fundamental criteria for stable coexistence to occur between two species is reciprocal invasibility when one of the species is at equilibrium (the resident) and the other species is rare. In this section we therefore analyse the impacts of spatial flows between two ecosystems on the equilibrium inorganic nutrient concentration in the presence of a single resident primary producer. We focus on the location of the net supply point because it is central to understand invasion conditions (equilibrium inorganic nutrient concentration) and thus coexistence in both locations. Stable coexistence between two species will be possible when each of them has a positive per capita growth rate when at low abundance (Chesson, 2000; Gravel et al. 2011). In terms of graphical representation of the dynamics, it means that the equilibrium nutrient concentration of each species when resident will be above the ZNGI of the invader. We study each spatial flow independently, even though they are likely to be colinear in nature, in order to distinguish their isolated effect on coexistence. We finish with the analysis of a special case where spatial flows and nutrient recycling promote unstable coexistence (alternate stable states) between two competing species. The analysis is essentially derived from a graphical interpretation of equilibrium vectors.

@@

Because we assume passive diffusion, the detritus will flow from the most productive location to the least productive one (Gravel et al., 2010a,b). The equilibrium total biomass of the producer at a location, and consequently the detritus biomass producer, is easily found on the two-dimensional space of nutrients (Fig. 2). The biomass will be proportional to the distance between the net supply point and the equilibrium inorganic nutrient concentration (in absence of diffusion). For instance, at Fig 2, the equilibrium producer and detritus biomass will be larger at the grey location (the most productive combinations of the two resources) than the black location (the least productive). Because it is passive, diffusion between the two locations has an averaging effect, meaning that it will tend to homogenize the detritus compartments from the two locations. Detritus will move from the most productive (grey) to the least productive (black) location. Consequently the net supply rate of both nutrients increases at the least productive location  while it decreases at the least productive location. The movement of the net supply point still follows however the projection of the consumption vector because the two nutrients are exchanged between the location exactly in the same proportion they are consumed (the quota of the detritus is the same as the quota of the producer). The equilibrium inorganic nutrient concentration thus remains unchanged by spatial flows of detritus, and consequently coexistence is neither promoted or impeded by detritus flows.

@@

%-----------------%
\begin{equation}
	\label{e:bnet}
	\Delta S_{jx}=q_{j}\phi r\widehat{D}_{x}e^{-1}
\end{equation}
%-----------------%

@@

which corresponds to the slope of the consumption vector (the ratio of quotas). Altough it does not influence stable coexistence, the spatial flow of detritus promotes unstable coexistence if the displacement of the net supply point from the original supply point is crossing the projection of an inferior competitor toward the coexistence area (see below), or alternatively it could lead to competitive exclusion if it moves in the opposite situation.

@@

@@

	This equation is already quite complicated and it is not possible to derive an instructive solution for the equilibrium nutrient concentration and producer densities in the two locations. This solution however simplifies considerably in the simpler, but still informative, situation where there is no nutrient recycling (i.e. $\phi = 1$ and $r = 0$):

@@

@@

It is easy to see the impact of source and sink dynamics on the equilibrium $N^{*}_{j}$ from this equation. The difference between producer densities will take a positive value for a source (by definition, $P_x>P_y$) and consequently $N^{*}_{j}$ will increase. At the opposite, $N^{*}_{j}$ will decrease for a sink where the difference is negative ($P_x<P_y$). Paradoxally, source and sink dynamics heterogenizes the ZNGI of a single species across the landscape (Fig 4), while it homogenizes the overall competitivity of the species at the regional scale \parencite{Mouquet2002}. It is thus difficult to see any general trend on coexistence, producer dispersal could either promote or prevent coexistence at the local scale. But eventually it will reduce coexistence at the regional scale because the most competitive species at the regional scale will exclude all other species (Mouquet and Loreau, 2002; Mouquet et al. 2006).

@@

@@

\section{Acknowledgments}
\newpage

%\section{References}
\printbibliography

\newpage

\section{Figures}

\begin{figure}[p]
   \centering
   \includegraphics[width=\textwidth]{R-Ratio_theory.pdf}
   \caption{Graphical interpretation of the resource-ratio theory for a single location. The panel A) schematizes the equilibrium state of the system with a single resident primary producer population. Nutrient cycling moves the location of the supply point, from location $S$ to the location $S’$. At equilibrium, the consumption vector is opposed to a net nutrient supply vector. The location of the equilibrium nutrient concentration will only be affected if the slope of the net supply vector is different of the slope of the consumption vector. Panel B) illustrates the conditions for coexistence. A second species could invade the system if its ZNGI is below the equilibrium nutrient concentration in the presence of the resident species. The equilibrium nutrient concentration would thus be found at the crossing of the two ZNGIs. Stable coexistence will occur if the location of the net supply points is found between the projection of the two consumption vectors (dotted lines) and if each species is more limited by the nutrient it requires the most (in other words, if the slope of the consumption vector increases as the ZNGI moves up and left).}
   \label{f:R-Ratio_theory}
\end{figure}
\newpage

\begin{figure}[p]
   \centering
   \includegraphics[width=0.45\textwidth]{DetritusDiffusion.pdf}
   \caption{Impacts of detritus diffusion on the equilibrium states of the metaecosystem with a single resident primary producer species. The filled symbols represent the original supply points (inorganic nutrient concentration in absence of consumption) while the open symbols represent the equilibrium quantities when accounting for consumption, recycling and diffusion. The figure shows that the net supply rate of both nutrients increases at the least productive location (in black) while it decreases at the least productive location (in grey). A liner functional response of the form $g_{ij}(N_{j}) = \beta_{ij}N_j$ was used. Parameters are: $e=0.1$, $r= 0.5$, $\phi=0.5$, $q_1 = 0.5$, $N^*_1=5$, $N^*_2=5$, $m =0.1$, $\beta_1 = mq_1/N^*_1$, $\beta_2 = mq_2/N^*_2$, $d_N = 0$, $d_P = 0$ and $d_D = 1$.
}
   \label{f:Detritus}
\end{figure}
\newpage

\begin{figure}[p]
   \centering
   \includegraphics[width=0.45\textwidth]{NutrientDiffusion.pdf}
   \caption{Impacts of inorganic nutrient dispersal on the equilibrium states of the metaecosystem with a single resident primary producer species. Same parameters as the Figure 2, except $r = 0.5$, $d_N = 0.1$, $d_P = 0$ and $d_D = 0$.}
   \label{f:Nutrients}
\end{figure}
\newpage

\begin{figure}[p]
   \centering
   \includegraphics[width=0.45\textwidth]{ProducerDiffusion.pdf}
   \caption{Impacts of primary producer dispersal on the equilibrium states of the metaecosystem with a single resident primary producer species. Same parameters as the Figure 3, except $d_N = 0$, $d_P = 10$ and $d_D = 0$.}
   \label{f:Producers}
\end{figure}
\newpage

\begin{figure}[p]
   \centering
   \includegraphics[width=0.45\textwidth]{AlternateStates.pdf}
   \caption{Detritus dispersal could potentially lead to alternate stable states, including locally stable coexistence and locally stable dominance of a single primary producer species. Parameters are: $e=0.1$, $r= 0.1$, $\phi=0.9$, $q_{11} = 1/3$, $N^*_{11}=5$, $N^*_{12}=10$, $m_1 =0.1$, $\beta_{11} = m_1q_{11}/N^*_{11}$, $\beta_{12} = m_1q_{12}/N^*_{12}$,$q_{21} = 2/3$, $N^*_{21}=10$, $N^*_{22}=5$, $m_2 =0.1$,$\beta_{21} = m_2q_{21}/N^*_{21}$, $\beta_{22} = m_2q_{22}/N^*_{22}$, $d_N = 0$, $d_P = 0$ and $d_D = 1$}
   \label{f:ASS}
\end{figure}
\newpage

>>>>

% TO DO
% schematic representation of the model
% definition of phi: lost during mineralization process (not an inorganic nutrient, to avoid confusion with eN)
% qij missing in Eq. 6
% eq 9: replace l by phi, and add qij, and replace N by P
% Localisation of the net supply point on Fig. 1
% Eq. 12: add the two coordinates in the text
% Vérifier la présence des chapeaux aux équations
% Explication modèle local: préciser que ça correspond aux équations sans les deltas
%

\documentclass[letterpaper,twocolumn,showkeys]{revtex4-1}
\usepackage[utf8]{inputenc}
\usepackage{color,dcolumn,graphicx,hyperref}

@@

\author{Dominique Gravel}
\author{Nicolas Mouquet}
\author{Mathew Leibold}
\author{François Massol}
\author{Tanguy Daufresne}

@@

\maketitle

@@

The interaction between community assembly (determining the species composition of local communities) and ecosystem attributes (determining the stocks and fluxes of nutrients and energy) is complex because it depends on so many distinct processes including nutrient consumption, recycling, export and import of materials by the biota.  The effects are even more complex because of stoichiometric relations that constrain how processes involving one element may be connected to another (e.g. nitrogen and phosphorous fluxes through an ecosystem is strongly linked in some processes such as plant uptake and not others such as recycling). Recent work on these issues has revealed important consequences of such stoichiometric mechanisms that operate within ecosystems (Sterner and Elser, Loreau).  However exchanges of materials and organisms between ecosystems may also be important in more spatially extended ‘meta-ecosystem’s (Loreau et al, Gravel et al 2010a, Gravel et al. 2010b) although their effects on stoichiometric mechanisms have not yet been well studied (but see Ryabov and Blasius 2011).

In the limit as exchanges between two ecosystems increase via enhanced diffusion there should be a pronounced homogenizing effects since the dynamics of the combined ecosystems will be identical to each subsystem if they are completely mixed.  However the effects at lower exchange rates are harder to predict since it may depend on which types of materials or organisms are the ones that move between ecosystems.  For example, if the stoichiometry of dispersing consumers, of dissolved available nutrients or of detritus are distinct, the stoichiometry of exchanges between ecosystems may be affected  in significantly different ways.

Here we use ‘niche based’ (Leibold 1995) approaches to resource competition between two consumers growing on two essential nutrient resources (Tilman 1982) in a metaecosystem consisting of two patches in source-sink situations (Gravel et al. 2010a).  We account for internal resource cycling within each patch (Daufresne and Hedin) and ask how the diffusive (unbiased, dependent on density or concentration differences between patches) exchanges of dissolved nutrients, detritus or consumers may differ in the consequences for consumer community composition and coexistence and for the concentrations of available nutrients.  We find that each type of exchange affects the process in fundamentally distinct ways. Diffusion affect the stoichiometry of the exchanges between patches due to variability in the stoichiometry of each type of material and on the indirect effects they have on resource competition.   Exchanges of detritus (assuming consumers and their detritus has a fixed stoichiometry) homogenize the productivity of the two patches without affecting any local stoichiometric effects, that exchanges of dissolved nutrients will homogenize both the productivity and stoichiometry of the two patches, and that exchanges of live consumers will have effects that maintain stoichiometry but affect nutrient depletion in patches.  We also show that the joint effects of detritus exchanges can affect coexistence and may generate alternate stable states in local communities.

@@

%-----------------%
\begin{equation}
	\label{e:bnet}
	\frac{dN_{jx}}{dt}=S_{jx}-eN_{jx}-\sum_{i=1}^{s}q_{ij}G_{ix}N_{jx}+\sum_{i=1}^{s}r(1-\phi)q_{ij}D_{ix}-\Delta_{Nx}
\end{equation}
%-----------------%

@@

The conditions for coexistence of primary producer species competing for two resources is best understood using a graphical representation of the model (Tilman, 1982). The fundamental criteria for stable coexistence to occur between two species is reciprocal invasibility when one of the species is at equilibrium (the resident) and the other species is rare. We first illustrate and derive tools to interpret the equilibrium state of the system with a single resident primary producer species and detritus mineralization at a single locality (the $\delta$ from Eqs 1,5 and 6 set to zero). We introduce dispersal for each compartment at the following section.

\begin{figure}[tb!]
   \centering
   \includegraphics[width=0.5\textwidth]{R-Ratio_theory.pdf}
   \caption{Graphical interpretation of the resource-ratio theory for a single location. The panel A) schematizes the equilibrium state of the system with a single resident primary producer population. Nutrient cycling moves the location of the supply point, from location $S$ to the location $S’$. At equilibrium, the consumption vector is opposed to a net nutrient supply vector. The location of the equilibrium nutrient concentration will only be affected if the slope of the net supply vector is different of the slope of the consumption vector. Panel B) illustrates the conditions for coexistence. A second species could invade the system if its ZNGI is below the equilibrium nutrient concentration in the presence of the resident species. The equilibrium nutrient concentration would thus be found at the crossing of the two ZNGIs. Stable coexistence will occur if the location of the net supply points is found between the projection of the two consumption vectors (dotted lines) and if each species is more limited by the nutrient it requires the most (in other words, if the slope of the consumption vector increases as the ZNGI moves up and left).}
   \label{f:R-Ratio_theory}
\end{figure}

@@

where $\overrightarrow{u'}_{x}$ is an altered nutrient supply vector. The consumption vector therefore points toward a net nutrient supply point $S'_{x}$, for which the coordinate of nutrient $j$ is is given by:

@@

 In this section we therefore analyse the impacts of spatial flows between two ecosystems on the equilibrium inorganic nutrient concentration in the presence of a single resident primary producer. We focus on the location of the net supply point because it is central to understand invasion conditions (equilibrium inorganic nutrient concentration) and thus coexistence in both locations. Stable coexistence between two species will be possible when each of them has a positive per capita growth rate when at low abundance (Chesson, 2000; Gravel et al. 2011). In terms of graphical representation of the dynamics, it means that the equilibrium nutrient concentration of each species when resident will be above the ZNGI of the invader. We study each spatial flow independently, even though they are likely to be colinear in nature, in order to distinguish their isolated effect on coexistence. We finish with the analysis of a special case where spatial flows and nutrient recycling promote unstable coexistence (alternate stable states) between two competing species. The analysis is essentially derived from a graphical interpretation of equilibrium vectors.

@@

\begin{figure}[tb]
   \centering
   \includegraphics[width=0.45\textwidth]{DetritusDiffusion.pdf}
   \caption{Impacts of detritus diffusion on the equilibrium states of the metaecosystem with a single resident primary producer species. The filled symbols represent the original supply points (inorganic nutrient concentration in absence of consumption) while the open symbols represent the equilibrium quantities when accounting for consumption, recycling and diffusion. The figure shows that the net supply rate of both nutrients increases at the least productive location (in black) while it decreases at the least productive location (in grey). A liner functional response of the form $g_{ij}(N_{j}) = \beta_{ij}N_j$ was used. Parameters are: $e=0.1$, $r= 0.5$, $\phi=0.5$, $q_1 = 0.5$, $N^*_1=5$, $N^*_2=5$, $m =0.1$, $\beta_1 = mq_1/N^*_1$, $\beta_2 = mq_2/N^*_2$, $d_N = 0$, $d_P = 0$ and $d_D = 1$.
}
   \label{f:Detritus}
\end{figure}

Because we assume passive diffusion, the detritus will flow from the most productive location to the least productive one (Gravel et al., 2010a,b). The equilibrium total biomass of the producer at a location, and consequently the detritus biomass producer, is easily found on the two-dimensional space of nutrients (Fig. 2). The biomass will be proportional to the distance between the net supply point and the equilibrium inorganic nutrient concentration (in absence of diffusion). 
% Give an explanation by refering to equation 6 (by isolating P)
% Will be proportional to the difference between S and eqN and to D,
% Where based on Eq. 5, D is proportional to P

For instance, at Fig 2, the equilibrium producer and detritus biomass will be larger at the grey location (the most productive combinations of the two resources) than the black location (the least productive). Because it is passive, diffusion between the two locations has an averaging effect, meaning that it will tend to homogenize the detritus compartments from the two locations. Detritus will move from the most productive (grey) to the least productive (black) location. Consequently the net supply rate of both nutrients increases at the least productive location  while it decreases at the least productive location. The movement of the net supply point still follows however the projection of the consumption vector because the two nutrients are exchanged between the location exactly in the same proportion they are consumed (the quota of the detritus is the same as the quota of the producer). The equilibrium inorganic nutrient concentration thus remains unchanged by spatial flows of detritus, and consequently coexistence is neither promoted or impeded by detritus flows.

@@

%-----------------%
\begin{equation}
	\label{e:bnet}
	\Delta S_{jx}=q_{j}\phi r\widehat{D}_{x}e^{-1}
\end{equation}
%-----------------%
% La dérivation du delta n'est pas super claire. Pourquoi le delta est disparu?
% Mentionner qu'il se retrouve dans la définition de Delta chapeau
% Le rendre explicite dans l'équation, comme c'est le cas pour les nutriments

@@

which corresponds to the slope of the consumption vector (the ratio of quotas). Altough it does not influence stable coexistence, the spatial flow of detritus promotes unstable coexistence if the displacement of the net supply point from the original supply point is crossing the projection of an inferior competitor toward the coexistence area (see below in the section "Alternate stable states"), or alternatively it could lead to competitive exclusion if it moves in the opposite situation. 
% Ce paragraphe finit la section de façon abrupte avec un lien rapide sur la coexistence
% Lien avec le cone de coexistence à la figure 1.
% Donne l'impression que la coexistence ne sera pas traitée davantage
% Anticiper un peu ce qui se trouve à la section suivante.
% Ou bien aller encore moins loin

@@

\begin{figure}[tb]
   \centering
   \includegraphics[width=0.45\textwidth]{NutrientDiffusion.pdf}
   \caption{Impacts of inorganic nutrient dispersal on the equilibrium states of the metaecosystem with a single resident primary producer species. Same parameters as the Figure 2, except $r = 0.5$, $d_N = 0.1$, $d_P = 0$ and $d_D = 0$.}
   \label{f:Nutrients}
\end{figure}

% Ajouter les vecteurs sur la figure, ajouter les flèches pour montrer les déplacements des supply points
% Les légendes sont violentes -- eg définition des N* dans le texte (au moment où on définit les ZNGI)

@@

	This equation is already quite complicated and it is not possible to derive an instructive solution for the equilibrium nutrient concentration and producer densities in the two locations. This solution however simplifies considerably in the simpler, but still informative, situation where there is no nutrient recycling (i.e. $\phi = 1$ or $r = 0$):

@@

\begin{figure}[tb]
   \centering
   \includegraphics[width=0.45\textwidth]{ProducerDiffusion.pdf}
   \caption{Impacts of primary producer dispersal on the equilibrium states of the metaecosystem with a single resident primary producer species. Same parameters as the Figure 3, except $d_N = 0$, $d_P = 10$ and $d_D = 0$.}
   \label{f:Producers}
\end{figure}

@@

It is easy to see the impact of source and sink dynamics on the equilibrium $N^{*}_{j}$ from this equation. The difference between producer densities will take a positive value for a source (by definition, $P_x>P_y$) and consequently $N^{*}_{j}$ will increase. At the opposite, $N^{*}_{j}$ will decrease for a sink where the difference is negative ($P_x<P_y$). Paradoxally, source and sink dynamics heterogenizes the ZNGI of a single species across the landscape (Fig 4), while it homogenizes the overall competitivity of the species at the regional scale (Mouquet and Loreau, 2002). It is thus difficult to see any general trend on coexistence, producer dispersal could either promote or prevent coexistence at the local scale. But eventually it will reduce coexistence at the regional scale because the most competitive species at the regional scale will exclude all other species (Mouquet and Loreau, 2002; Mouquet et al. 2006).

@@

\begin{figure}[tb]
   \centering
   \includegraphics[width=0.45\textwidth]{AlternateStates.pdf}
   \caption{Detritus dispersal could potentially lead to alternate stable states, including locally stable coexistence and locally stable dominance of a single primary producer species. Parameters are: $e=0.1$, $r= 0.1$, $\phi=0.9$, $q_{11} = 1/3$, $N^*_{11}=5$, $N^*_{12}=10$, $m_1 =0.1$, $\beta_{11} = m_1q_{11}/N^*_{11}$, $\beta_{12} = m_1q_{12}/N^*_{12}$,$q_{21} = 2/3$, $N^*_{21}=10$, $N^*_{22}=5$, $m_2 =0.1$,$\beta_{21} = m_2q_{21}/N^*_{21}$, $\beta_{22} = m_2q_{22}/N^*_{22}$, $d_N = 0$, $d_P = 0$ and $d_D = 1$}
   \label{f:ASS}
\end{figure}

@@

Metaecosystems could be very complicated.

Here we show noentheless that mechanisms remains fairly simple
Not really new coexistence mechanism, but still ....

\section{Acknowledgments}
